The {\em Diff} MILP model and PCA for \\ {\em real-time} DNN robustness certification

### 2025-11-15 CAV 2026/introduction.tex
INTRO

	\item We develop a novel {\em diff MILP encoding} for the global robustness problem, %called the {\em "2v" model}, 
	where the variables are the values of the perturbed neurons, as well as the difference between the original and the perturbed neuron values (called the {\em diff variables}, introduced in \cite{diff}). We study and encode how the {\em diff variables} evolve after passing through a ReLU (Prop.~\ref{Prop2}). Compared with the {\em classical MILP model} \cite{MILP} employed in \cite{vhagar,lipshitz,ITNE}, which considers the input and the perturbation but not their difference ({\em diff variable}), we keep the same number of 2 binary variables per ReLU. Experimentally however, 
	%the results are more accurate for the same runtime.
	%the linear relaxation is much more accurate, as each {\em diff variable} can be bounded after a ReLU as a function of the value of the {\em diff variables} before the ReLU, whereas the linear relaxations of the classical model is extremely inaccurate, as the variables are independent of each other. This was observed in \cite{lipshitz,ITNE}, and constraints encoding
	%a part of the linear relaxation of our "2v" model were added explicitly. 
	%Experimentally, 
	our {\em diff MILP model} provides more accurate bounds at a given runtime
	(Table \ref{table.classical}).
	%The number of variables a priori doubles compared with local robustness, from each neuron value in the perturbed image to each neuron value in the perturbed image {\em and} in the original image, as the original image is no more fixed. 
	%Recall that the worst case complexity of MILP is exponential in the number of binary variables \cite{DivideAndSlide}. 
	%A straightforward MILP model would be to use the  for each of these variables, as in \cite{vhagar,lipshitz}. The main issue with the classical model is that its linear relaxations is extremely inaccurate, as the variables are independent of each other. Instead, we develop another exact MILP model, 
	\item Further from the {\em diff MILP model}, which is exact, 
	we develop two abstract MILP models, which are more efficient but also possibly less accurate.
	Namely, the "1b+2" model replaces the 2 binary variables by 1 binary (for the diff variable) and 2 linear variables (linear relaxation for the perturbation); while the "1b" model has a unique binary variable per ReLU, only considering the {\em diff variables}. 
	%\item We adapt the Solution Aware Scoring from \cite{ATVA25} to our novel MILP models, in order to select the most important ReLUs to be treated using complex binary variables, while less important variables are treated using linear relaxation. The chosen number of binary variables depends upon the complexity of the DNN as well as the targeted runtime.

@@

\item Experimentally, the exact {\em diff MILP model} computes bounds 
$\beta_{C,D}$ which are 1 order of magnitude better than with the classical MILP encoding, and much closer (Table \ref{table.classical}) to the lower bound than with variants of the classical encoding; namely Vhagar \cite{vhagar} which reduces the number of binary variables in a complete fashion; and ITNE \cite{ITNE} which adds linear relaxation for the diff variables.
Our "1b+2" variant offers good trade-offs, reaching better bounds in the same runtime as the diff MILP model. The "1b" model converges faster but bounds are looser, interesting when runtime is very limited or problem is very complex (Tables \ref{table.mnist},\ref{table.pipe}).
Overall, the "1b+2" model obtains bounds $\beta_{C,D}$ which enables the robustness certification of $69\%$ of MNIST images for an $L_1$-perturbation of $1$, resp. $89\%$ for $.5$. The online process adds only
0.0005s of latency per image, and 2000 images/second can be treated per CPU core.
\end{enumerate}

>>>>

	\item We develop a novel {\em diff MILP encoding} for the global robustness problem, %called the {\em "2v" model}, 
	where the variables are the values of the perturbed neurons, as well as the difference between the original and the perturbed neuron values (called the {\em diff variables}, introduced in \cite{diff}). We study and encode how the {\em diff variables} evolve after passing through a ReLU (Prop.~\ref{Prop2}). Compared with the {\em classical MILP model} \cite{MILP} employed in \cite{vhagar,lipshitz,ITNE}, which considers the input and the perturbation but not their difference ({\em diff variable}), we keep the same number of 2 binary variables per ReLU. 
	%the results are more accurate for the same runtime.
	%the linear relaxation is much more accurate, as each {\em diff variable} can be bounded after a ReLU as a function of the value of the {\em diff variables} before the ReLU, whereas the linear relaxations of the classical model is extremely inaccurate, as the variables are independent of each other. This was observed in \cite{lipshitz,ITNE}, and constraints encoding
	%a part of the linear relaxation of our "2v" model were added explicitly. 
	%Experimentally, 
	However, our {\em diff MILP model} is more efficient, one reason being 
	the accuracy of its linear relaxation.
	%The number of variables a priori doubles compared with local robustness, from each neuron value in the perturbed image to each neuron value in the perturbed image {\em and} in the original image, as the original image is no more fixed. 
	%Recall that the worst case complexity of MILP is exponential in the number of binary variables \cite{DivideAndSlide}. 
	%A straightforward MILP model would be to use the  for each of these variables, as in \cite{vhagar,lipshitz}. The main issue with the classical model is that its linear relaxations is extremely inaccurate, as the variables are independent of each other. Instead, we develop another exact MILP model, 
	\item Further from the {\em diff MILP model}, which is exact, 
	we develop two abstract MILP models, which are more efficient but also possibly less accurate.
	Namely, the "1b+2" model replaces the 2 binary variables by 1 binary (for the diff variable) and 2 linear variables (linear relaxation for the perturbation); while the "1b" model has a unique binary variable per ReLU, only considering the {\em diff variables}. 
	%\item We adapt the Solution Aware Scoring from \cite{ATVA25} to our novel MILP models, in order to select the most important ReLUs to be treated using complex binary variables, while less important variables are treated using linear relaxation. The chosen number of binary variables depends upon the complexity of the DNN as well as the targeted runtime.

@@

\item Experimentally, the exact {\em diff MILP model} computes upper bounds 
which are 1 order of magnitude better than with the classical MILP encoding, and much closer (Table \ref{table.classical}) to the lower bounds than with variants of the classical encoding; namely reducing the number of binary variables in a complete fashion (Vhagar \cite{vhagar}); or adding linear relaxation for the diff variables (ITNE \cite{ITNE}).
Our "1b+2" variant offers good trade-offs, reaching better bounds in the same runtime as the diff MILP model. The "1b" model converges faster but bounds are looser, interesting when runtime is very limited or problem is very complex (Tables \ref{table.mnist},\ref{table.pipe}).
Further, using PCA reduces $\beta_{C,D}$ by up to an order of magnitude.
Overall, the "1b+2" model using PCA enables the robustness certification of $69\%$ of MNIST images for an $L_1$-perturbation of $1$, resp. $89\%$ for $.5$. The online process adds only 0.0005s of latency per image, and 2000 images/second can be treated per CPU core.
\end{enumerate}
